Transferability of clinical prediction models for early trauma care

### 2019-07-30 ManuscriptMarkdown.Rmd
Update ManuscriptMarkdown.Rmd

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "July 29, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Introduction
Trauma, defined as an injury to living tissue by an extrinsic agent [@RN63], 
results in approximately 3,8 million deaths annually across the globe [@RN1; @RN21]. 
Each year 9% of the world’s deaths are because of trauma, with the leading causes being 
road traffic accidents, suicide and homicide. Predictions indicate the incidence of all 
these causes are likely to increase by 2030 [@RN42].

@@

## Setting
In Sweden, trauma is the most common cause of death before 44 years of age. Up 
to 10% of all major trauma patients suffer some form of disability. 
The societal effects of trauma are also substantial, with the loss
of work days comparable to those lost from cardiovascular and malignant diseases
combined [@RN23].

@@

## Participants 
The eligibility criteria were adult patients aged 15 years or above registered in SweTrau 
between 2011 and 2016. The study aimed to assess adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. The Swedish guidelines for trauma activation define children as age <15
[@RN90].

@@

![](fig1.jpg)  
**Fig. 1: Sample development in the high and low volume centre data set.** Illustrates 
creation of development and validation sample in the high and low volume data set. 
Initial split based on number of patients. Temporal split made using date of traumatic event.

@@

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates. In all data sets, model transfer between samples resulted in
substantial uncertainty in its effect on model performance. When comparing
transferred performance to the reciveing sample validation performance the
resulting differance also showed considerable uncertainty.

The most notable effects on model performance following model transfer were seen
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increase of the mistriage rate by 
`r substr(format(highest.comp.mistriage), 1, 5)`.  Mainly contributing to this was
a marked increase in overtriage.  For this transfer the 95% UI ranged from 
`r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating a high level of uncertainty and
that our findings are compatible with a `r UI.top.bot[1]` decrease to a 
`r UI.top.bot[2]` increase in mistriage. The negative 95% UI of `r UI.top.bot[1]`
would signifiy an improvement in model performance. The higher 95% UI of 
`r UI.top.bot[2]` would mean a marked increase in model mistriage, compared to the
point estimate of `r substr(format(highest.comp.mistriage), 1, 5)`.

@@

External validation as the evaluation of a clinical prediction model in a
setting in which it was not originally developed has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when being externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62], showed models
significantly underestimating the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
suggests caution when transferring models.

@@

### Limitations
One limitation is that originally all single centres with a valid number of 
events were to constitute a data set, but only one centre provided a sufficient 
number of events. Leading to the loss of the data set: Individual centres. 
This data set might have provided results that could reflect realistic 
situations as major trauma centres regularly exchange data.

Furthermore, this study used GCS, RR and SBP as predictors for 30-day
mortality. Additional predictors such as age, mechanism
of injury, or newer predictors as shock index might increase model
performance. Furthermore, we used 30-day mortality as outcome, recognizing
that late mortality, time in hospital, functional impairment, and morbidity are
also important outcome factors in the context of trauma care.

The predetermined analysis plan was to be followed in a step-by-step fashion, to
avoiding confirmation bias. Programming challenges and late-time code inclusions
required a flexible approach to this analysis plan, sometimes with revisions of
previous steps. This is not optimal, and efforts were made to not deviate in any
major way. Using GitHub, all code was made public for inspection and to ensure
reproducibility [@RN59].

## Clinical applications
This study shows that mistriage can potentially increase following model
transfer, and more prominently in certain cases. This supports the view that
model transfers should be done with caution, as performance might
deteriorate.

## Conclusion
Model transfer resulted in large and varying levels of uncertainty in its effect
on mistriage. Therefore model transfer can be unpredictable, potentially leading 
to increased mistriage, but in some cases also to a decrease in mistriage. 
Noticeably, transfer of the `r colnames(highest.comp.mistriage)` model lead to an 
increased mistriage of `r highest.comp.mistriage`. This observed effect was also 
associated with considerable uncertainty. Recognizing the limitations of this study, 
the authors believe further studies are warranted dueto the potential economic 
costs and patient consequences of poor model transfers.

>>>>

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "July 31, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Introduction
Trauma, defined as an injury to living tissue by an extrinsic agent [@RN63], 
results in approximatly 458,5 million cases anually across the globe [@RN999], 
and an estimated 3,8 million deaths [@RN1; @RN21]. Each year 9% of the world’s 
deaths are because of trauma, with the leading causes being road traffic accidents, 
suicide and homicide. Predictions indicate the incidence of these causes are 
likely to increase by 2030 [@RN42].

@@

## Setting
In Sweden, trauma is the most common cause of death before 44 years of age, and 
10% of major trauma patients suffer some form of disability. 
The societal effects of trauma are substantial, with the loss
of work days comparable to those lost from cardiovascular and malignant diseases
combined [@RN23].

@@

## Participants 
The eligibility criteria were patients aged 15 years or above registered in SweTrau 
between 2011 and 2016. The study aimed to assess adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. The Swedish guidelines for trauma activation define children as age <15
[@RN90].

@@

![](fig1.jpg)  
**Fig. 1: Sample development in the high and low volume centre data set.**  Initial 
split based on number of patients. Temporal split made using date of traumatic event.

@@

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates. In all data sets, model transfer between samples resulted in
substantial uncertainty in its effect on model performance. When comparing
transferred performance to the receiving sample validation performance, the
resulting differences also showed considerable uncertainty.

The most notable effects on model performance following model transfer were seen
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increase of the mistriage rate by 
`r substr(format(highest.comp.mistriage), 1, 5)`.  Mainly contributing to this was
a marked increase in overtriage.  For this transfer, the 95% UI ranged from 
`r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating a high level of uncertainty and
that our findings are compatible with a `r UI.top.bot[1]` decrease to a 
`r UI.top.bot[2]` increase in mistriage. The negative 95% UI of `r UI.top.bot[1]`
would signify an improvement in model performance. The higher 95% UI of 
`r UI.top.bot[2]` would mean a marked increase in model mistriage, compared to the
point estimate of `r substr(format(highest.comp.mistriage), 1, 5)`.

Due to centralized trauma care, non-metropolitan centres generelly have fewer 
resources, and less experience reciving trauma patients. They may rely 
on clinical prediction models more heavily for accurate trauma triage, and an 
increased mistriage leads to further strain on already limited resources
due to increased overtriage. Increased undertriage could also possibly lead to
increased patient mortality

@@

External validation as the evaluation of a clinical prediction model in a
setting in which it was not originally developed has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when being externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62], showed models
significantly underestimating the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
would suggest caution when transferring models.

@@

### Limitations
Originally, all single centres with a valid number of 
events were to constitute a data set, but only one centre provided a sufficient 
number of events, leading to the loss of the individual centres data set. 
This data set could have provided interesting results reflecting realistic 
situations.

This study used GCS, RR and SBP as predictors for 30-day
mortality. Additional predictors such as age, mechanism
of injury, or newer predictors like shock index might increase model
performance. Furthermore, we used 30-day mortality as outcome, recognizing
that late mortality, time in hospital, functional impairment, and morbidity are
also important outcome factors in the context of trauma care.

The predetermined analysis plan was to be followed in a step-by-step fashion. 
Programming challenges and late-time code inclusions required a flexible approach 
to this analysis plan, sometimes with revisions of previous steps. This is not 
optimal, and efforts were made to not deviate in any major way. Using GitHub, 
all code was made public for inspection and to ensure reproducibility [@RN59].

## Conclusion
Model transfer resulted in large and varying levels of uncertainty in its effect
on mistriage. Therefore, model transfer can be unpredictable, potentially leading 
to increased mistriage, but in some cases also to a decrease in mistriage. 
Noticeably, transfer of the `r colnames(highest.comp.mistriage)` model lead to an 
increased mistriage of `r highest.comp.mistriage`. This observed effect was also 
associated with considerable uncertainty. Recognizing the limitations of this study, 
the authors believe further studies are warranted due to the potential economic 
costs and patient consequences of poor model transfers.
